---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/"
)
```

# RSoptSC 

```{r, echo = FALSE, out.width = '60%'}
knitr::include_graphics(path = "man/figures/logo.svg")
```

RSoptSC is an R package for SoptSC: performing analysis and prediction on single-cell RNA sequencing (scRNA-seq) data. SoptSC can infer cell-cell communication between single cells, and completes multiple single-cell analysis tasks from a single coherent framework. This enables the reconstruction of complex lineage relationships alongside prediction of specific feedback or feedforward interactions. 

* Demo SoptSC at `vignette("Joost_et_al")`

* Read the SoptSC preprint on [bioRxiv](https://www.biorxiv.org/content/early/2018/05/12/168922)


## Installation

```{r, eval = FALSE, echo = TRUE}
install.packages("devtools")
library(devtools)
install_github("mkarikom/RSoptSC")
```


## Features

* Regularized network inference on single-cell gene expression.

* Unsupervised clustering

* Semi-supervised or unsupervised pseudotime by root cell/cluster inference and graph embedding 
