---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/"
)
```

# RSoptSC <img src="man/figures/logo.svg" align="right" alt="" width="320" />

[![Build Status](https://travis-ci.com/mkarikom/RSoptSC.svg?branch=master)](https://travis-ci.com/mkarikom/RSoptSC)

## Updates

* Demo SoptSC at [`vignette("Joost_et_al")`](https://mkarikom.github.io/RSoptSC/articles/Joost_et_al.html)

* Read the SoptSC preprint on [bioRxiv](https://www.biorxiv.org/content/10.1101/168922v3)

* SoptSC is also available as a MATLAB package [here](https://github.com/wangshuxiong/soptsc)


## Installation

```{r, eval = FALSE, echo = TRUE}
install.packages("devtools")
library(devtools)
install_github("mkarikom/RSoptSC")
```


## Features

* Inference of cell-cell communication between single cells

* Integration of multiple analyses in a consistent mathematical framework: clustering, marker genes, pseudotime, and lineage inference

* Cell-cell similarity matrix construction to improve clustering

* NMF-based marker gene identification

* Prediction of the number of clusters present (via eigengap properties of the similarity matrix)

* Prediction of the initial cell in pseudotime


## Documentation

Full details of RSoptSC and examples are available [here](https://mkarikom.github.io/RSoptSC).
